---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

# ncdfCF

<!-- badges: start -->

[![Lifecycle: Experimental](https://img.shields.io/badge/Lifecycle-Experimental-red.svg)](https://lifecycle.r-lib.org/articles/stages.html) [![CRAN Status](https://www.r-pkg.org/badges/version/ncdfCF)](https://cran.r-project.org/package=ncdfCF) [![CRAN Downloads](https://cranlogs.r-pkg.org/badges/grand-total/ncdfCF)](https://cran.r-project.org/package=ncdfCF) [![License: GPL v3](https://img.shields.io/badge/License-MIT-blue.svg)](https://mit-license.org) [![Last commit](https://img.shields.io/github/last-commit/pvanlaake/ncdfCF)](https://github.com/pvanlaake/ncdfCF/commits/main)
[![R-CMD-check](https://github.com/pvanlaake/ncdfCF/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pvanlaake/ncdfCF/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `ncdfCF` package provides an easy to use interface to netCDF resources in R,
either in local files or remotely on a THREDDS server. It is built on the 
`RNetCDF` package which, like package `ncdf4`, provides a basic interface to the 
`netcdf` library, but which lacks an intuitive user interface. Package `ncdfCF` 
provides a high-level interface using functions and methods that are familiar to 
the R user. It reads the structural metadata and also the attributes upon 
opening the resource. In the process, the `ncdfCF` package also applies CF 
Metadata Conventions to interpret the data. This currently applies to:

 * The **axis designation**. The three mechanisms to identify the axis each 
 dimension represents are applied until an axis is determined.
 * The **time dimension**. Time is usually encoded as an offset from a datum.
 Using the `CFtime` package these offsets can be turned into intelligible dates
 and times, for all 9 defined calendars.
 * **Bounds** information. When present, bounds are read and used in analyses.
 * **Discrete dimensions**, optionally with character labels. When labels are
 provided, these will be used as `dimnames` for the axis. (Note that this also
 applies to generic numeric axes with labels defined.)
 * **Parametric vertical coordinates** are read, including variables listed in
 the `formula_terms` attribute.
 * **Auxiliary coordinates** are identified and read. This applies to both 
 **scalar axes** and **auxiliary longitude-latitude grids**. Data on non-Cartesian 
 grids can be automatically rectified to a longitude-latitude grid if an auxiliary
 grid is present in the resource.
 * The **grid_mapping** variables, providing the coordinate reference system 
 (CRS) of the data, with support for all defined objects in the latest EPSG 
 database as well as "manual" construction of CRSs.

##### Basic usage

Opening and inspecting the contents of a netCDF resource is very 
straightforward:

```{r basic_example}
library(ncdfCF)

# Get any netCDF file
fn <- system.file("extdata", "ERA5land_Rwanda_20160101.nc", package = "ncdfCF")

# Open the file, all metadata is read
(ds <- open_ncdf(fn))

# ...or very brief details
names(ds)
dimnames(ds)

# Variables and dimensions can be accessed through standard list-type extraction syntax
(t2m <- ds[["t2m"]])

ds[["longitude"]]

# Regular base R operations simplify life further
dimnames(ds[["pev"]]) # A variable: list of dimension names
dimnames(ds[["longitude"]]) # A dimension: vector of dimension element values

# Access attributes
ds[["pev"]]$attribute("long_name")
```

If you just want to inspect what data is included in the netCDF resource, use
the `peek_ncdf()` function: 

```{r peek}
(ds <- peek_ncdf(fn))
```

##### Extracting data

There are three ways to read data for a variable from the resource:

 * **`data():`** The `data()` method returns all data of a variable, including 
 its metadata, in a `CFData` instance.
 * **`[]`:** The usual R array operator gives you access to the raw, non-interpreted
 data in the netCDF resource. This uses index values into the
 dimensions and requires you to know the order in which the dimensions are 
 specified for the variable. With a bit of tinkering and some helper functions
 in `ncdfCF` this is still very easy to do.
 * **`subset()`:** The `subset()` method lets you specify what you want to
 extract from each dimension in real-world coordinates and timestamps, in 
 whichever order. This can also rectify non-Cartesian grids to regular
 longitude-latitude grids.
 
```{r extract}
# Extract a timeseries for a specific location
ts <- t2m[5, 4, ]
str(ts)

# Extract the full spatial extent for one time step
ts <- t2m[, , 12]
str(ts)
```

Note that the results contain degenerate dimensions (of length 1). This by 
design when using basic `[]` data access because it allows attributes to be 
attached in a consistent manner. When using the `subset()` method, the data is 
returned as an instance of `CFData`, including axes and attributes:

```{r subset}
# Extract a specific region, full time dimension
(ts <- t2m$subset(list(X = 29:30, Y = -1:-2)))

# Extract specific time slices for a specific region
# Note that the dimensions are specified out of order and using alternative
# specifications: only the extreme values are used.
(ts <- t2m$subset(list(T = c("2016-01-01 09:00", "2016-01-01 15:00"),
                       X = c(29.6, 28.8),
                       Y = seq(-2, -1, by = 0.05))))
```

The latter two methods will read only as much data from the netCDF resource as 
is requested.

##### Exporting and saving data

A `CFData` object can be exported to a `data.table` or to a `terra::SpatRaster`
(3D) or `terra::SpatRasterDataset` (4D) for further processing. Obviously, these 
packages need to be installed to utilise these methods.

```{r export}
# install.packages("data.table")
library(data.table)
head(dt <- ts$data.table())

#install.packages("terra")
suppressMessages(library(terra))
(r <- ts$terra())
terra::plot(r[names(r)[1]])
```

A `CFData` object can also be written back to a netCDF file. The object will
have all its relevant attributes and properties written together with the actual 
data: axes, bounds, attributes, CRS.

```{r save, eval=FALSE}
# Save a CFData instance to a netCDF file on disk
ts$save("~/path/file.nc")
```

##### A note on Discrete Sampling Geometries

Discrete Sampling Geometries (DSG) map almost directly to the venerable 
`data.frame` in R (with several exceptions). In that sense, they are rather
distinct from array-based data sets. At the moment there is no specific code for
DSG, but the simplest layouts can currently already be read (without any 
warranty). Various methods, such as `CFVariable::subset()` or `CFData::array()` 
will fail miserably, and you are well-advised to try no more than the empty 
array indexing operator `CFVariable::[]` which will yield the full data variable  
with column and row names set as an array, of `CFVariable::data()` to get the 
whole data variable as a `CFData` object for further processing. You can 
identify a DSG data set by the `featureType` attribute of the `CFDataset`.

More comprehensive support for DSG is in the development plan.

## Development plan

Package `ncdfCF` is in the early phases of development. It supports reading of 
groups, variables, dimensions, user-defined data types, attributes and data from 
netCDF resources in "classic" and "netcdf4" formats; and can write single data
variables back to a netCDF file. From the CF Metadata 
Conventions it supports identification of dimension axes, interpretation of the 
"time" dimension, name resolution when using groups, reading of "bounds" 
information, parametric vertical coordinates, auxiliary coordinate variables,
labels, and grid mapping information.

Development plans for the near future focus on supporting the below features:

##### netCDF

 * Support for writing of data sets.

##### CF Metadata Conventions

 * Calculate parametric vertical coordinates.
 * Aggregation, using the CFA convention.
 * Support for discrete sampling geometries.
 * Interface to "standard_name" libraries and other "defined vocabularies".
 * Compliance with CMIP5 / CMIP6 requirements.

## Installation

**`r colorize("CAUTION:", "red")`** Package `ncdfCF` is still in the early 
phases of development. While extensively tested on multiple well-structured 
datasets, errors may still occur, particularly in datasets that do not adhere to 
the CF Metadata Conventions.

Installation from CRAN of the latest release:

```
install.packages("ncdfCF")
```

You can install the development version of `ncdfCF` from [GitHub](https://github.com/) with:

```
# install.packages("devtools")
devtools::install_github("pvanlaake/ncdfCF")
```
