---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ncdfCF

<!-- badges: start -->

[![Lifecycle: Experimental](https://img.shields.io/badge/Lifecycle-Experimental-red.svg)](https://lifecycle.r-lib.org/articles/stages.html) [![License: GPL v3](https://img.shields.io/badge/License-MIT-blue.svg)](https://mit-license.org) [![Last commit](https://img.shields.io/github/last-commit/pvanlaake/ncdfCF)](https://github.com/pvanlaake/ncdfCF/commits/main)
<!-- badges: end -->

The `ncdfCF` package provides an easy to use interface to NetCDF resources in R, 
either in local files or remotely on a THREDDS server.

It is built on the `RNetCDF` package which provides a basic interface to the
NetCDF library, but which lacks an intuitive user experience - effectively it is
a transliteration of the `netcdf` library API into R and some tweaks to adjust
to the R environment (e.g. data types and internal representation). Package 
`ncdf4` does one better by performing the tedious task of reading the structural
metadata from the resource that is needed for a basic understanding of the 
contents, such as dimension and variable details, but the essential library API 
concept remains with functions that directly map to the `netcdf` library 
functions. One would really need to understand the NetCDF data model and 
implementation details to effectively use these packages. (For instance, most 
data describing a dimension is stored as a variable! So to read the `dimnames()` 
of a dimension you'd have to call `var.get.nc()` or `ncvar_get()`.) Neither 
package loads the attributes of the dimensions, variables and the dataset 
("global" variables), which is essential to *understand* what the dimensions and 
variables represent.

Package `ncdfCF` provides a high-level interface using functions and methods 
that are familiar to the R user. It reads the structural metadata and also the
attributes upon opening the resource. In the process, the `ncdfCF` package also 
applies CF Metadata Conventions to interpret the data. This currently applies 
to:

 * The **axis designation**. The three mechanisms to identify the axis each 
 dimension represents are applied until an axis is determined.
 * The **time dimension**. Time is usually encoded as an offset from a datum.
 Using the `CFtime` package these offsets can be turned into intelligible dates
 and times, for all 9 defined calendars.
 * **Bounds** information. When present, bounds are read and used in analyses.

##### Basic usage

Opening and inspecting the contents of a NetCDF resource is very 
straightforward:

```{r basic_example}
library(ncdfCF)

# Get any NetCDF file
fn <- system.file("extdata", "ERA5land_Rwanda_20160101.nc", package = "ncdfCF")

# Open the file, all metadata is read
ds <- ncdfDataset(fn)

# Easy access in understandable format to all the details
ds

# Variables can be accessed through standard list-type extraction syntax
t2m <- ds[["t2m"]]
t2m

# Same with dimensions, but now without first putting the object in a variable
ds[["longitude"]]

# Regular base R operations simplify life further
dimnames(ds[["pev"]]) # A variable: list of dimension names
dimnames(ds[["longitude"]]) # A dimension: vector of dimension element values

# Access attributes
attribute(ds[["pev"]], "long_name")
```

##### Extracting data

One of the perpetual headaches of users of NetCDF files is to extract the data.
If you want to get all the data for a variable then neither `RNetCDF` nor 
`ncdf4` are particularly troublesome:

``` r
library(ncdf4)
nc <- nc_open(fn)
vars <- names(nc$vars)
d <- ncvar_get(nc, vars[[1]])
```

But what if you are interested in only a small area or a month of data while the 
resource has global data spanning multiple years? In both `RNetCDF` and `ncdf4`
packages you'd have to work out how your real-world boundaries translate to 
indices into the variable array of interest and then populate `start` and 
`count` arguments to pass on to `var.get.nc()` or `ncvar_get()`. Many R users
default to simply reading the entire array and then extracting the area of
interest using standard R tools. That is wasteful at best (lots of I/O, RAM 
usage, CPU cycles) and practically impossible with some larger NetCDF resources 
that have variables upwards of 1GB in size.

Enter `ncdfCF`. With `ncdfCF` you have two options to extract data for a 
variable:

 * **`[]`**: Using R's standard extraction operator `[` you work directly with
 the index values into the array dimensions: `d <- t2m[3:5, 1:4, 1:10]`. You can
 leave out dimensions to extract everything from that dimension (but you have to
 indicate the position, just like in regular arrays). So to get the first 5 
 "time" slices from `t2m`: `d <- t2m[, , 1:5]`. Not specifying anything gets you 
 the whole array: `d <- t2m[]`. This works for any number of dimensions, you 
 simply have to adjust the number of positions that you specify. You still need 
 to know the indices into
 the arrays but `ncdfCF` has some helper functions to get you those.
 * **`subset()`**: The `subset()` method is more flexible than `[]` because it
 requires less knowledge of how the data in the variable is structured,
 particularly the order of the dimensions. While many NetCDF resources "behave" in
 their dimension order, there is no guarantee. With `subset()` you supply a 
 list with items for each dimension (by axis or dimension name, in any order)
 and each item containing a vector of real-world coordinates to extract. As an 
 example, to extract values of a variable `x` for Australia for the year 2020 you call
 `subset(x, list(X = 112:154, Y = -9:-44, T = c("2020-01-01", "2021-01-01")))`.

Both approaches lend themselves well to the `apply()` family of functions for
processing. Importantly, these functions are to access data from the NetCDF 
resource so you can tweak the size of your request to the capacity of the 
computer, without exhausting RAM.

## Development plan

Package `ncdfCF` is in the early phases of development. It supports reading of 
dimensions, variables, attributes and data from NetCDF resources in "classic" and
"NetCDF4" formats. From the CF Metadata Conventions it supports identification
of dimension axes, interpretation of the "time" dimension, and reading of 
"bounds" information.

Development plans for the near future focus on supporting the below features:

##### NetCDF

 * Support for writing.
 * Support for "group" information in "NetCDF4" formatted resources.

##### CF Metadata Conventions

 * Full support for discrete or categorical dimensions.
 * Interface to "standard_name" libraries and other "defined vocabularies".
 * Compliance with CMIP5 / CMIP6 requirements.

## Installation

**CAUTION:** Package `ncdfCF` is still in the early phases of development.
While extensively tested on multiple well-structured datasets, errors may still
occur, particularly in datasets that do not adhere to the CF Metadata 
Conventions.

Package `ncdfCF` has not yet been submitted to CRAN.

You can install the development version of `ncdfCF` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pvanlaake/ncdfCF")
```
