---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ncdfCF

<!-- badges: start -->

[![Lifecycle: Experimental](https://img.shields.io/badge/Lifecycle-Experimental-green.svg)](https://lifecycle.r-lib.org/articles/stages.html) [![CRAN Status](https://www.r-pkg.org/badges/version/ncdfCF)](https://cran.r-project.org/package=ncdfCF) [![CRAN Downloads](https://cranlogs.r-pkg.org/badges/grand-total/ncdfCF)](https://cran.r-project.org/package=ncdfCF) [![License: GPL v3](https://img.shields.io/badge/License-MIT-blue.svg)](https://mit-license.org) [![Last commit](https://img.shields.io/github/last-commit/R-CF/ncdfCF)](https://github.com/R-CF/ncdfCF/commits/main)
[![R-CMD-check](https://github.com/R-CF/ncdfCF/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-CF/ncdfCF/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `ncdfCF` package provides an easy to use interface to netCDF resources in R, either in local files or remotely on a THREDDS server. It is built on the `RNetCDF` package which, like package `ncdf4`, provides a basic interface to the `netcdf` library, but which lacks an intuitive user interface. Package `ncdfCF` provides a high-level interface using functions and methods that are familiar to the R user. It reads the structural metadata and also the attributes upon opening the resource. In the process, the `ncdfCF` package also applies CF Metadata Conventions to interpret the data. This currently applies to:

 * The **axis designation**. The three mechanisms to identify the axis each dimension represents are applied until an axis is determined.
 * The **time coordinate**. Time is usually encoded as an offset from an origin. Using the `CFtime` package these offsets can be turned into intelligible dates and times, for all defined calendars.
 * **Bounds** information. When present, bounds are read and used in analyses.
 * **Discrete coordinates**, optionally with character labels. When labels are provided, these will be used as `dimnames` for the axis. (Note that this also applies to generic numeric axes with labels defined.)
 * **Parametric vertical coordinates** are computed using the `formula_terms` attribute, for two ocean formulations.
 * **Auxiliary coordinates** are identified and read. This applies also to **scalar axes** and **auxiliary longitude-latitude grids**. Auxiliary coordinates can be activated by the user and then used in display, selection and processing. Data on non-Cartesian grids can be automatically rectified to a longitude-latitude grid if an auxiliary grid is present in the resource.
 * The **cell measure variables** are read and linked to any data variables referencing them. Cell measure variables that are external to the netCDF resource with the referring data variable can be linked to the data set and then they are immediately available to the referring data variables.
 * **Labels**, as separate variables identified through the `coordinates` attribute of axes, are read, including when multiple sets of labels are defined for a single axis. Users can select which set of labels to make active for display, selection and processing.
 * The **grid_mapping** variables, providing the coordinate reference system (CRS) of the data, with support for all defined objects in the latest EPSG database as well as "manual" construction of CRSs.

## Basic usage

Opening and inspecting the contents of a netCDF resource is very straightforward:

```{r basic_example}
library(ncdfCF)

# Get any netCDF file
fn <- system.file("extdata", "ERA5land_Rwanda_20160101.nc", package = "ncdfCF")

# Open the file, all metadata is read
(ds <- open_ncdf(fn))

# ...or very brief details
ds$var_names
ds$axis_names

# Variables and axes can be accessed through standard list-type extraction syntax
(t2m <- ds[["t2m"]])

ds[["longitude"]]

# Regular base R operations simplify life further
dimnames(ds[["pev"]]) # A variable: list of axis names
dimnames(ds[["longitude"]]) # An axis: vector of axis coordinate values

# Access attributes
ds[["pev"]]$attribute("long_name")
```

If you just want to inspect what data is included in the netCDF resource, use the `peek_ncdf()` function: 

```{r peek}
peek_ncdf(fn)
```

## Extracting data

There are three ways to read data for a data variable from the resource: 

 * **`[]`:** The usual R array operator gives you access to the raw, non-interpreted data in the netCDF resource. This uses index values into the dimensions and requires you to know the order in which the dimensions are specified for the variable. With a bit of tinkering and some helper functions in `ncdfCF` this is still very easy to do.
 * **`subset()`:** The `subset()` method lets you specify what you want to extract from each dimension in real-world coordinates and timestamps, in whichever order. This can also rectify non-Cartesian grids to regular longitude-latitude grids.
 * **`profile()`:** Extract "profiles" from the data variable. This can take different forms, such as a temporal or depth profile for a single location, but it could also be a zonal field (such as a transect in latitude - atmospheric depth for a given longitude) or some other profile in the physical space of the data variable.
 
```{r extract}
# Extract a timeseries for a specific location - see also the `profile()` method
ts <- t2m[5, 4, ]
str(ts)

# Extract the full spatial extent for one time step
ts <- t2m[, , 12]
str(ts)
```

Note that the results contain degenerate dimensions (of length 1). This by design when using basic `[]` data access because it allows attributes to be attached in a consistent manner. When using the `subset()` method, the data is returned as an instance of `CFVariable`, including axes and attributes:

```{r subset}
# Extract a specific region, full time dimension
(ts <- t2m$subset(list(X = 29:30, Y = -1:-2)))

# Extract specific time slices for a specific region
# Note that the dimensions are specified out of order and using alternative
# specifications: only the extreme values are used.
(ts <- t2m$subset(list(T = c("2016-01-01 09:00", "2016-01-01 15:00"),
                       X = c(29.6, 28.8),
                       Y = seq(-2, -1, by = 0.05))))
```

The latter two methods will read only as much data from the netCDF resource as is requested.

##### Make a profile of data

It is often useful to extract a "profile" of data for a given location or zone, such as a timeseries of data. The `profile()` method has some flexible options to support this:

 - Profile specific locations, with multiple locations specified per call, returning the data as a (set of) `CFVariable` instance(s) or as a single `data.table`.
 - Profile zones, such as a latitude band or an atmospheric level. Data is returned as a new `CFVariable` instance(s).
 
In all cases, you can profile over any of the axes and over any number of axes.

Note that the `profile()` method returns data for the grid cells closest to the specified location. That is different from the `subset()` method, which will return data as it is recorded in the netCDF resource.

```{r profile}
rwa <- t2m$profile(longitude = c(30.07, 30.07, 29.74), latitude = c(-1.94, -1.58, -2.60), 
                   .names = c("Kigali", "Byumba", "Butare"), .as_table = TRUE)
head(rwa)
attr(rwa, "value")
```

Some critical metadata is recorded in the "value" attribute: original long name and the physical unit.

When you provide coordinates for all axes but one, you get a profile of values along the remaining axis, as shown above. If you provide fewer axis coordinates you get progressively higher-order results. To get a latitudinal transect, for instance, provide only a longitude coordinate:

```{r transect}
(trans29_74 <- t2m$profile(longitude = 29.74, .names = "lon_29_74"))
```

Note that there is only a single longitude coordinate left, at exactly the specified longitude.

## Summarising data over time

With the `summarise()` method you can apply a function over the data to generate summaries. You could, for instance, summarise daily data to monthly means. These methods use the specific calendar of the "time" axis. The return value is a new `CFVariable` object.

```{r summarise}
# Summarising hourly temperature data to calculate the daily maximum temperature
t2m$summarise("tmax", max, "day")
```

A function may also return a vector of multiple values, in which case a list is returned with a new `CFVariable` object for each return value of the function. This allows you to calculate multiple results with a single call. You could write your own function to tailor the calculations to your needs. Rather than just calculating the daily maximum, you could get the daily maximum, minimum and diurnal range in one go:

```{r summarise_multi}
# Function to calculate multiple daily stats
# It is good practice to include a `na.rm` argument in all your functions
daily_stats <- function(x, na.rm = TRUE) {
  # x is the vector of values for one day
  minmax <- range(x, na.rm = na.rm)
  diurnal <- minmax[2L] - minmax[1L]
  c(minmax, diurnal)
}

# Call summarise() with your own function
# The `name` argument should have as many names as the function returns results
(stats <- t2m$summarise(c("tmin", "tmax", "diurnal_range"), daily_stats, "day"))
```

Note that you may have to update some attributes after calling `summarise()`. You can use the `set_attribute()` method on the `CFVariable` objects to do that.

## Create new netCDF objects

You can convert a suitable R object into a `CFVariable` instance quite easily. R objects that are supported include arrays, matrices and vectors of type logical, integer, numeric or logical.

```{r create_basic}
arr <- array(rnorm(120), dim = c(6, 5, 4))
as_CF("my_first_CF_object", arr)
```

Usable but not very impressive. The axes have dull names without any meaning and the coordinates are just a sequence along the axis.

If the R object has `dimnames` set, these will be used to create more informed axes. More interestingly, if your array represents some spatial data you can give your `dimnames` appropriate names ("lat", "lon", "latitude", "longitude", case-insensitive) and the corresponding axis will be created (if the coordinate values in the `dimnames` are within the domain of the axis type). For "time" coordinates, these are automatically detected irrespective of the name.

```{r create_dimnames}
# Note the use of named dimnames here - these will become the names of the axes
dimnames(arr) <- list(lat = c(45, 44, 43, 42, 41, 40), lon = c(0, 1, 2, 3, 4), 
                      time = c("2025-07-01", "2025-07-02", "2025-07-03", "2025-07-04"))

(obj <- as_CF("a_better_CF_object", arr))

# Axes are of a specific type and have basic attributes set
obj$axes[["lat"]]

obj$axes[["time"]]
```

You can further modify the resulting `CFVariable` by setting other properties, such as attributes or a coordinate reference system. Once the object is complete, you can export or save it.

## Exporting and saving data

A `CFVariable` object can be exported to a `data.table` or to a `terra::SpatRaster` (3D) or `terra::SpatRasterDataset` (4D) for further processing. Obviously, these packages need to be installed to utilise these methods.

```{r export}
# install.packages("data.table")
library(data.table)
head(dt <- ts$data.table())

#install.packages("terra")
suppressMessages(library(terra))
(r <- stats[["diurnal_range"]]$terra())
terra::plot(r)
```

A `CFVariable` object can also be written back to a netCDF file. The object will have all its relevant attributes and properties written together with the actual data: axes, bounds, attributes, CRS. The netCDF file is of version "netcdf4" and will have the axes oriented in such a way that the file has maximum portability (specifically, data will be stored in row-major order with increasing Y values).

```{r save, eval=FALSE}
# Save a CFVariable instance to a netCDF file on disk
stats[["diurnal_range"]]$save("~/path/file.nc")
```

##### A note on Discrete Sampling Geometries

Discrete Sampling Geometries (DSG) map almost directly to the venerable `data.frame` in R (with several exceptions). In that sense, they are rather distinct from array-based data sets. At the moment there is no specific code for DSG, but the simplest layouts can currently already be read (without any warranty). Various methods, such as `CFVariable::subset()` or `CFVariable::array()` will fail miserably, and you are well-advised to try no more than the empty array indexing operator `CFVariable::[]` which will yield the full data variable with column and row names set as an array, of `CFVariable::`data.table` for a format that matches the structure of a typical table closest. You can identify a DSG data set by the `featureType` attribute of the `CFDataset`.

More comprehensive support for DSG is in the development plan.

## Development plan

Package `ncdfCF` is still being developed. It supports reading of all data objects from netCDF resources in "classic" and "netcdf4" formats; and can write single data arrays back to a netCDF file. From the CF Metadata Conventions it supports identification of axes, interpretation of the "time" axis, name resolution when using groups, reading of grid cell boundary information, auxiliary coordinate variables, labels, cell measures, attributes and grid mapping information, among others.

Development plans for the near future focus on supporting the below features:

##### netCDF

 * Support for writing of complex data sets (single `CFVariable` instances can already be written to file).

##### CF Metadata Conventions

 * Cell methods
 * Aggregation, using the CFA convention.
 * Support for discrete sampling geometries.
 * Compliance with CMIP5 / CMIP6 requirements.

## Installation

Package `ncdfCF` is still being developed. While extensively tested on multiple well-structured data sets, errors may still occur, particularly in data sets that do not adhere to the CF Metadata Conventions. The API may still change and although care is taken not to make breaking changes, sometimes this is unavoidable.

Installation from CRAN of the latest release:

```
install.packages("ncdfCF")
```

You can install the development version of `ncdfCF` from [GitHub](https://github.com/) with:

```
# install.packages("devtools")
devtools::install_github("R-CF/ncdfCF")
```
