---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

# ncdfCF

<!-- badges: start -->

[![Lifecycle: Experimental](https://img.shields.io/badge/Lifecycle-Experimental-red.svg)](https://lifecycle.r-lib.org/articles/stages.html) [![License: GPL v3](https://img.shields.io/badge/License-MIT-blue.svg)](https://mit-license.org) [![Last commit](https://img.shields.io/github/last-commit/pvanlaake/ncdfCF)](https://github.com/pvanlaake/ncdfCF/commits/main)
<!-- badges: end -->

The `ncdfCF` package provides an easy to use interface to NetCDF resources in R, 
either in local files or remotely on a THREDDS server. It is built on the 
`RNetCDF` package which, like package `ncdf4`, provides a basic interface to the 
`netcdf` library, but which lacks an intuitive user interface. Package `ncdfCF` 
provides a high-level interface using functions and methods that are familiar to 
the R user. It reads the structural metadata and also the attributes upon 
opening the resource. In the process, the `ncdfCF` package also applies CF 
Metadata Conventions to interpret the data. This currently applies to:

 * The **axis designation**. The three mechanisms to identify the axis each 
 dimension represents are applied until an axis is determined.
 * The **time dimension**. Time is usually encoded as an offset from a datum.
 Using the `CFtime` package these offsets can be turned into intelligible dates
 and times, for all 9 defined calendars.
 * **Bounds** information. When present, bounds are read and used in analyses.

##### Basic usage

Opening and inspecting the contents of a NetCDF resource is very 
straightforward:

```{r basic_example}
library(ncdfCF)

# Get any NetCDF file
fn <- system.file("extdata", "ERA5land_Rwanda_20160101.nc", package = "ncdfCF")

# Open the file, all metadata is read
ds <- open_ncdf(fn)

# Easy access in understandable format to all the details
ds

# Variables can be accessed through standard list-type extraction syntax
t2m <- ds[["t2m"]]
t2m

# Same with dimensions, but now without first putting the object in a variable
ds[["longitude"]]

# Regular base R operations simplify life further
dimnames(ds[["pev"]]) # A variable: list of dimension names
dimnames(ds[["longitude"]]) # A dimension: vector of dimension element values

# Access attributes
attribute(ds[["pev"]], "long_name")
```

##### Extracting data

There are two ways to read data for a variable from the resource:

 * **`[]`:** The usual R array operator. This uses index values into the
 dimensions and requires you to know the order in which the dimensions are 
 specified for the variable. With a bit of tinkering and some helper functions
 in `ncdfCF` this is still very easy to do.
 * **`subset()`:** The `subset()` method lets you specify what you want to
 extract from each dimension in real-world coordinates and timestamps, in 
 whichever order.
 
```{r extract}
# Extract a timeseries for a specific location
ts <- t2m[5, 4, ]
str(ts)

# Extract the full spatial extent for one time step
ts <- t2m[, , 12]
str(ts)
```

Note that the results contain degenerate dimensions (of length 1). This by 
design because it allows attributes to be attached in a consistent manner.

```{r subset}
# Extract a specific region, full time dimension
ts <- subset(t2m, list(X = 29:30, Y = -1:-2))
str(ts)

# Extract specific time slices for a specific region
# Note that the dimensions are specified out of order and using alternative
# specifications: only the extreme values are used.
ts <- subset(t2m, list(T = c("2016-01-01 09:00", "2016-01-01 15:00"),
                       X = c(29.6, 28.8),
                       Y = seq(-2, -1, by = 0.05)))
str(ts)
```

Both of these methods will read data from the NetCDF resource, but only as much
as is requested. 

## Development plan

Package `ncdfCF` is in the early phases of development. It supports reading of 
dimensions, variables, attributes and data from NetCDF resources in "classic" and
"NetCDF4" formats. From the CF Metadata Conventions it supports identification
of dimension axes, interpretation of the "time" dimension, and reading of 
"bounds" information.

Development plans for the near future focus on supporting the below features:

##### NetCDF

 * Support for writing.
 * Support for "group" information in "NetCDF4" formatted resources.

##### CF Metadata Conventions

 * Full support for discrete or categorical dimensions.
 * Interface to "standard_name" libraries and other "defined vocabularies".
 * Compliance with CMIP5 / CMIP6 requirements.

## Installation

**`r colorize("CAUTION:", "red")`** Package `ncdfCF` is still in the early 
phases of development. While extensively tested on multiple well-structured 
datasets, errors may still occur, particularly in datasets that do not adhere to 
the CF Metadata Conventions.

Package `ncdfCF` has not yet been submitted to CRAN.

You can install the development version of `ncdfCF` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pvanlaake/ncdfCF")
```
