---
title: "Example 1. How to open and visualise Copernicus Marine Data in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 1. How to open and visualise Copernicus Marine Data in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Copernicus Marine Service has a [page](https://help.marine.copernicus.eu/en/articles/5040824-how-to-open-and-visualize-copernicus-marine-data-in-r) that demonstrates how to open and visualise Copernicus marine data in `R` using the `ncdf4` package. This page shows how to get the same results - and more - using the `ncdfCF` package.

## Main packages

This example uses the `ncdfCF` package to access and interpret the marine data. For visualisation some optional packages are loaded later on.

## Open netCDF files and load the data
```{r open}
# Open the netCDF file
library(ncdfCF)
(ds <- open_ncdf("../../tests/testdata/cmems_mod_med_phy-tem_anfc_4.2km_P1M-m_1748594717636.nc"))
```

All overall information about the file is listed:

  - Variables: `bottomT` and `thetao`, with 3 and 4 axes, respectively.
  
  - Global attributes: title, source, references, etc.

Also important is the global attribute `Conventions`, which tells us that this netCDF file is encoded using the CF-1.11 metadata conventions. The `ncdfCF` package uses this to interpret all the information in the file that give the variables "meaning", such as their axes, units, time coordinates, etc.

(Please note that the warnings at the top are generated because there is an error in the definition of the "coordinates" attribute of the data variables. That has no negative effect on using this file. Further down you'll see what caused these warnings to be generated.)

## Get an ocean variable

There are two variables in the data set: `bottomT` and `thetao.` We'll work here with `thetao`: Sea temperature.

```{r thetao}
# Assign the theta data variable
(thetao <- ds[["thetao"]])
```

(Coming back to the warnings at the beginning: The "coordinates" attribute has the value "time depth lat lon". If you look at the axes, however, you can see that the first two are called "longitude" and "latitude", rather than their short versions "lon" and "lat": a warning will be generated.)

From looking at the properties of the data set we already knew that data variable thetao has 4 axes. Now that we drill down to the details of the data variable you can see the range of values and their physical unit. You can also inspect the coordinates along the axes quite easily:

```{r coordinates}
# The "time" coordinates
thetao$axes[["time"]]$coordinates

# The "depth" coordinates
thetao$axes[["depth"]]$coordinates
```

The `thetao` data variable has 12 months along its "time" axis and 141 "depth" values. Let's look at the temperature at the surface. The first coordinate value is 1.018237m below the surface but it actually represents the surface layer. We can extract this data using the `$subset()` method of the data variable, using a margin around the value:

```{r surface}
(surfaceT <- thetao$subset(depth = c(0, 2)))
```

Note that there is only 1 "depth" value left. The axis is still there so that you can look at its properties.

## Quick map plot

```{r image_plot}
# Get the longitude and latitude coordinates
longitude <- thetao$axes[["longitude"]]$coordinates
latitude <- thetao$axes[["latitude"]]$coordinates

# Make a map of the first time slice: January
library(RColorBrewer)
image(longitude, latitude, surfaceT$raw()[,,,1], col = rev(brewer.pal(10, "RdBu")))
```

The `lattice` package's `levelplot()` function enables you to obtain a better map:

```{r lattice_plot}
library(lattice)
grid <- expand.grid(lon=longitude, lat=latitude)
cutpts <- c(12,13,14,15,16,17,18,19,20)
levelplot(surfaceT$raw()[,,,1] ~ lon * lat,
          data=grid, region=TRUE,
          pretty=T, at=cutpts, cuts=9,
          col.regions=(rev(brewer.pal(9,"RdBu"))), contour=0,
          xlab = "Longitude", ylab = "Latitude",
          main = "Sea Water Potential Temperature (°C)"
)
```

## Make a depth profile of temperature

You can make a temperature profile for a location and time (or a transect) very easily. Here we get the temperate profile at increasing depths for the month of July in the middle of the Tyrrhenian Sea, one of the deeper parts of the Mediterranean.

```{r profile}
# Make a vertical profile of temperature at a single location for the month of July 2024
tprof <- thetao$profile(longitude = 12.71, latitude = 40.33, time = "2024-07-01", .names = "Tyrrhenian_Sea")

# Turn the CFVariable object into a data.table
library(data.table)
Tjuly <- tprof$data.table()
head(Tjuly)
```

This table can be used for further analysis, but you can also make very nice graph with the `ggplot2` package:

```{r profile_plot}
library(ggplot2)
ggplot(Tjuly) + geom_line(aes(depth, Tyrrhenian_Sea)) + theme_bw() + xlim(0, 4000) + 
  labs(x = "Depth (m)", y = "Temperature (°C)", title =  "Tyrrhenian Sea", subtitle = "Location: 12.71E - 40.33N  Month: July 2024")
```

The temperature drops off very rapidly in the first 100m, then slowly decreases to its lowest level at the bottom of the sea.

(The warning is harmless: Recall that the 141 depth values go down to almost 6,000m. Here the seafloor is at about 3,600m and the remaining depth values are set to NaN (not a number). These can obviously not be plotted.)
